Layered Grammar for Alluvial Plots

### 2020-01-01 paper.rmd
typos in paper

**Repeated ordinal measures data:**
Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination. Ordinal variables are most appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well. A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order. In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.

@@

**Signed categorical data:**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exeters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, reflected across the dimensional axis. Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign. Future use cases may demonstrate the practical utility of this functionality.

@@

I am grateful to many users for their feedback on every version of this package. Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant from the National Institute of Dental and Craniofacial Research (5T90DE021989-07).

>>>>

**Repeated ordinal measures data:**
Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size, though some follow user-provided hierarchies, and most insert gaps between categories for easy visual discrimination. Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well. A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order. In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.

@@

**Signed categorical data:**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot, reflected across the dimensional axis. Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent definitionally signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even if cases change sign. Future use cases may demonstrate the practical utility of this functionality.

@@

I am grateful to many users for their feedback on every version of this package. Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
